---
title: "Clinical Trials with a Time-to-Event Endpoint: Final group challenge"
author: "Kaspar Rufibach"
date: "Last change: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_document:
    highlight: pygments
    number_sections: yes
    self_contained: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# Getting started{-}

- Download the raw R quarto file [here]().
- Start up Rstudio. Then open the R quarto exercise file.
- Make sure that you have rpact version $\geq 3.5.1$ installed.
- Load the rpact package by executing the code below in R.
- Start answering the questions below by adding your own R code. Use the rpact vignettes [here](https://rpact.org/vignettes/planning/rpact_survival_planning_example/)  and [here](https://rpact.org/vignettes/planning/rpact_survival_examples/) for guidance.

$\Rightarrow$ **Good Luck!**

# Load rpact package {-}
```{r, include=TRUE, echo=TRUE}
# Load rpact
library(rpact)
packageVersion("rpact") 
library(knitr)
```


# Trial with time-to-event endpoint

## The trial

A phase III, randomized, multicenter, open-label, two-arm study designed to evaluate treatment A versus B. 

The **general assumptions for the sample size assessment** are:

- Randomization in a 2:1 ratio.
- The dropout rate is 5% for arm A and 10% for arm B.
- The recruitment of approximately 250 patients will take place over 5 months. 

The following additional assumptions are made:

- Exponentially distributed PFS in the control arm with a median of 4 months.
- Median PFS improvement vs. control of 3.3 months (medians 7.3 vs. 4 months, i.e. a HR of approximately 0.55). 
- Log-rank test at a two-sided significance level of 0.05, power 80%.

Based on these assumptions, compute the following quantities:

- Number of events required to reach the targeted power.
- The MDD.
- The clinical cutoff date for the primary analysis.

## Questions and solutions

*a) Verify the sample size calculation.*

**Solution:**
```{r, include=TRUE, echo=TRUE}
sampleSize <- getSampleSizeSurvival(sided = 2,alpha = 0.05, beta = 0.2, 
    allocationRatioPlanned = 2,    # 2:1 randomization
    lambda2 = getLambdaByMedian(4), lambda1 = getLambdaByMedian(7.3),
    dropoutRate1 = 0.05, dropoutRate2 = 0.10, dropoutTime = 12,
    accrualTime = 0:5, 
    accrualIntensity = rep(50, 5),
    maxNumberOfSubjects = 250)
summary(sampleSize)
```

*b) How many events would be expected to occur until exactly 2 and 4 months, respectively, from first patient randomized? (Use function `getEventProbabilities`.)*

**Solution:**
```{r, include=TRUE, echo=TRUE}
# Probability of an event until 2 months and 4 months  
prob <- getEventProbabilities(time = c(2, 4), 
    allocationRatioPlanned = 2,    # 2:1 randomization
    lambda2 = getLambdaByMedian(4), lambda1 = getLambdaByMedian(7.3),
    dropoutRate1 = 0.05, dropoutRate2 = 0.10, dropoutTime = 12,
    accrualTime = 0:5,
    accrualIntensity = rep(50, 5),
    maxNumberOfSubjects = 250)
prob

# Expected number of events (for n = 250)
250 * prob$overallEventProbabilities
```
